### 2023-01-24 article.tex
Data is plural

%
% 'Data are ...', i.e. use plural
%

@@

Here we define paradata as the data that describes the process of generating data.
In genetic epidemiology, the data generated is mostly the results 
of an analysis (e.g. predicting a person having a disease),
as done by computer code.
In such context, one could argue incorrectly that
it is the -usually English- scientific paper that best describes which computation steps 
have taken place.
However, this has the unrealistic 
assumption that there is a perfect match between the paper and the code.
In this chapter it is argued that the source code should should be supplied,
as this is the true paradata: if the paper and code disagree, it is the
code that has generated the results.
The chapter concludes by some rules how to better code to serve as paradata,
and hence allowing computational research to be truly reproducible.

@@

In this paper, the definition of paradata
is 'information about how data was generated',
as shown schematically in the left panel of Figure \ref{fig:figure_1}.
Paradata does not have a clear definition yet (hence this book). 
Common elements in the definitions of paradata are 
(1) paradata are usually not the primary data, (2) paradata describe
a process, (3) paradata describe the way of collecting primary data, or 
(4) paradata describe the way the primary data is generated.

@@

A simple example would be the data at eBird (\url{https://ebird.org/}),
where people can submit wher a bird of a certain species was spotted.
The name of the person that spotted that bird is paradata, 
as it is that person that generated the data and determined of
which species the bird was, 
as shown schematically in the middle panel of Figure \ref{fig:figure_1}.

@@

Paradata is relevant, as it allows primary data to be re-used.
In the eBird example, it helps assess the trustworthiness of the 
data, as beginner birdwatchers tend to make more mistakes in identifying
the correct bird species, as compared to experts.
By using the name and/or experience of the birder as a covariate,
we can re-use the dataset in other contexts as well,
as now this example can be generalized to
a classification problem of unlabeled
data for learning agents with different amounts 
of experience.

@@

This paper uses genetic epidemiology as a specific example,
to illustrate that the computer code is paradata,
that this paradata is relevant, as it helps 
to assess the correctness of the results,
where the correctness of the results is relevant for 
science in general and healthcare in particular.
However, any field that uses computation in its experiments
could be used as an example.

@@

To \emph{obtain} the raw genetic data, there may have been fieldwork and/or lab-work
involved. For example, a researcher needs to go into the field to take
samples (e.g. blood) of humans or other species. 
To analyse these samples,
a researcher needs a lab. 
Depending on the analysis, there may be a big bioinformatics step
to aggregate the measurements into usable data. 
Although this paper does not focus on the code run to 
aggragate raw and noisy measurements into usable data, the same arguments
can be made for that code as in the example: that code is 
the ground truth of how the raw data is converted into useful genetic data.

@@

Programming code (from now on: 'code') is paradata, 
as it is data that describes how data is generated.
Code is data that is usually in the form of text, 
spread over one or more files, that describes an experiment/annalysis.
The computational experiment (from now on: 'experiment') collects the data
and then looks for patterns.
Those patterns (again, a type of data) are the results of an experiment.

The (scientific) article is metadata, as it is data that describes other data:
an article describes the
experiment; an article describes what the code does in a natural 
language, such as English.
However, English (or any other natural language) is not the best
candidate to describe how the data is collected,
as it only has a loose connection with collecting the data.
If the code and article of an experiment disagree,
it is the code that actually let the data be collected.  
Instead, an article is metadata about a research.

@@

Hosted code has many additional features, such as
a version-controlled history, continuous integration scripts
and automatically generated metrics.
A code hosting website produces and shares this metadata.
This metadata can help to determine the
reproducibility, correctness and quality of code
of computational experiments.
Below, those features and why these are important are discussed.

@@

Containers allow a computation experiment to be highly reproducible:
given the same data, an experiment put into a container will give
the same results on different platforms, at least in theory.
In practice, differences may be observed when peripheral factors
are different, such as the random numbers as generated by an operating
system, or data that is downloaded from additional online sources.

@@

For a computational experiment to be reproducible,
an independent person should be able to
reproduce the same results.
The reproduce the same results, one needs the same
input data and the same code (and the same 
dependencies and environment).
When the input data is sensitive, however, and thus cannot be shared with
an independent researcher, he/she thus cannot reproduce the
experiment.

@@

For research to be reproducible, one ideally has access to
both the data used and the code.
In some fields, such as genetic epidemiology, the data is
sensitive, hence cannot be released,
yet there are methods being devised to run code on sensitive
data with assured privacy \cite{zhang2016review,azencott2018machine}.
Additionally, it follows software development's best practices 
(especially the practice to test code automatically) 
to release a simulated/public dataset, with
the additional benefit that this dataset can be used for comparisons.

@@

\begin{table}[h]
  \begin{tabular}{lp{5cm}p{5cm}}
    Term      & Definition                                            & Example                                                         \\
    \hline
    Code      & Body of text to be run by a computer                  & The scripts run by a computational experiment                   \\
    Data      & Individual facts, statistics, or items of information & A SNP that has a significant association                        \\
    Genotype  & The DNA allele at a certain location                  & AA, AC, CG, GT, ...                                             \\
    Paradata  & Data that describes the process of generating data    & The code to conclude that a SNP has a significant association   \\
    Phenotype & How an organism looks like in the broadest sense      & The concentration of IL6RA in the blood                         \\
    Primary data & Data that is worked with in a research             & The genotype of individuals   \\
    Metadata  & Data that provides information about other data       & The article that describes an experiment                        \\
    Raw data  & Data directly from the real world, that is not yet ready to be primary data & The raw results of getting the genotypes of individuals   \\
    Trait     & A phenotype                                           & The concentration of IL6RA in the blood                         
  \end{tabular}
  \caption{Terms used in this paper, and their definitions}
  \label{tab:definitions}
\end{table}

>>>>

@@

Here we define paradata as the data that describe the process of generating data.
In genetic epidemiology, the data generated are mostly the results 
of an analysis (e.g. predicting a person having a disease),
as done by computer code.
In such context, one could argue incorrectly that
it is the -usually English- scientific paper that best describes which computation steps 
have taken place.
However, this has the unrealistic 
assumption that there is a perfect match between the paper and the code.
In this chapter it is argued that the source code should should be supplied,
as these are the true paradata: if the paper and code disagree, it is the
code that has generated the results.
The chapter concludes by some rules how to better code to serve as paradata,
and hence allowing computational research to be truly reproducible.

@@

In this paper, the definition of paradata
is 'information about how data were generated',
as shown schematically in the left panel of Figure \ref{fig:figure_1}.
The term 'paradata' does not have a clear definition yet (hence this book). 
Common elements in the definitions of paradata are 
(1) paradata are usually not the primary data, (2) paradata describe
a process, (3) paradata describe the way of collecting primary data, or 
(4) paradata describe the way the primary data are generated.

@@

A simple example would be data at eBird (\url{https://ebird.org/}),
where people can submit wher a bird of a certain species was spotted.
The name of the person that spotted that bird is paradata, 
as it is that person that generated the data and determined of
which species the bird was, 
as shown schematically in the middle panel of Figure \ref{fig:figure_1}.

@@

Paradata are relevant, as it allows primary data to be re-used.
In the eBird example, it helps assess the trustworthiness of the 
data, as beginner birdwatchers tend to make more mistakes in identifying
the correct bird species, as compared to experts.
By using the name and/or experience of the birder as a covariate,
we can re-use the dataset in other contexts as well,
as now this example can be generalized to
a classification problem of unlabeled
data for learning agents with different amounts 
of experience.

@@

This paper uses genetic epidemiology as a specific example,
to illustrate that the computer code is paradata,
that this type of paradata is relevant, as it helps 
to assess the correctness of the results,
where the correctness of the results is relevant for 
science in general and healthcare in particular.
However, any field that uses computation in its experiments
could be used as an example.

@@

To \emph{obtain} the raw genetic data, there may have been fieldwork and/or lab-work
involved. For example, a researcher needs to go into the field to take
samples (e.g. blood) of humans or other species. 
To analyse these samples,
a researcher needs a lab. 
Depending on the analysis, there may be a big bioinformatics step
to aggregate the measurements into usable data. 
Although this paper does not focus on the code run to 
aggragate raw and noisy measurements into usable data, the same arguments
can be made for that code as in the example: that code is 
the ground truth of how the raw data are converted into useful genetic data.

@@

Programming code (from now on: 'code') is paradata, 
as it is data that describes how data are generated.
Code is data that is usually in the form of text, 
spread over one or more files, that describes an experiment/annalysis.
The computational experiment (from now on: 'experiment') collects the data
and then looks for patterns.
Those patterns (again, a type of data) are the results of an experiment.

The (scientific) article is metadata, as it is data that describes other data:
an article describes the
experiment; an article describes what the code does in a natural 
language, such as English.
However, English (or any other natural language) is not the best
candidate to describe how the data are collected,
as it only has a loose connection with collecting the data.
If the code and article of an experiment disagree,
it is the code that actually let the data be collected.  
Instead, an article is metadata about a research.

@@

Hosted code has many additional features, such as
a version-controlled history, continuous integration scripts
and automatically generated metrics.
A code hosting website produces and shares these metadata.
These metadata can help to determine the
reproducibility, correctness and quality of code
of computational experiments.
Below, those features and why these are important are discussed.

@@

Containers allow a computation experiment to be highly reproducible:
given the same data, an experiment put into a container will give
the same results on different platforms, at least in theory.
In practice, differences may be observed when peripheral factors
are different, such as the random numbers as generated by an operating
system, or data that are downloaded from additional online sources.

@@

For a computational experiment to be reproducible,
an independent person should be able to
reproduce the same results.
The reproduce the same results, one needs the same
input data and the same code (and the same 
dependencies and environment).
When the input data are sensitive, however, and thus cannot be shared with
an independent researcher, one thus cannot reproduce the
experiment.

@@

For research to be reproducible, one ideally has access to
both the data used and the code.
In some fields, such as genetic epidemiology, the data are
sensitive, hence cannot be released,
yet there are methods being devised to run code on sensitive
data with assured privacy \cite{zhang2016review,azencott2018machine}.
Additionally, it follows software development's best practices 
(especially the practice to test code automatically) 
to release a simulated/public dataset, with
the additional benefit that this dataset can be used for comparisons.

@@

\begin{table}[h]
  \begin{tabular}{lp{5cm}p{5cm}}
    Term      & Definition                                            & Example                                                         \\
    \hline
    Code      & Body of text to be run by a computer                  & The scripts run by a computational experiment                   \\
    Data      & Individual facts, statistics, or items of information & A SNP that has a significant association                        \\
    Genotype  & The DNA allele at a certain location                  & AA, AC, CG, GT, ...                                             \\
    Paradata  & Data that describe the process of generating data     & The code to conclude that a SNP has a significant association   \\
    Phenotype & How an organism looks like in the broadest sense      & The concentration of IL6RA in the blood                         \\
    Primary data & Data that are worked with in a research            & The genotype of individuals   \\
    Metadata  & Data that provide information about other data        & The article that describes an experiment                        \\
    Raw data  & Data directly from the real world, that are not yet ready to be primary data & The raw results of getting the genotypes of individuals   \\
    Trait     & A phenotype                                           & The concentration of IL6RA in the blood                         
  \end{tabular}
  \caption{Terms used in this paper, and their definitions}
  \label{tab:definitions}
\end{table}
